Deciphering the Relationship Betweeen Race, Income, and Political Support in the 2020 United States Election

### 2024-03-17 paper/paper.qmd
abstract and introduction done

---
title: "Deciphering the Relationship Betweeen Race, Income, and Political Support in the 2020 United States Election"
author: 
  - Liban Timir
thanks: "Code and data are available at: https://github.com/libant/US-Political-Support-Analysis"
date: 03-12-2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In the contemporary political landscape of the United States, marked by increasing polarization, understanding the foundations of political support has never been more crucial. The Cooperative Election Study (CES) of 2020 presents a dataset to explore this phenomenon. This dataset not only captures a wide array of sociopolitical attitudes and behaviors but also provides insights into the demographic and socioeconomic factors that influence political support. This offers a valuable lens through which to examine the complex dynamics of political engagement.

This study aims to fill a critical gap in the existing literature on political support in the United States by leveraging the CES 2020 data to examine the role of sociodemographic factors in shaping political preferences and behaviors. Previous research, while extensive, has often focused on more generalized measures of political engagement or on specific demographic groups, without fully addressing the multifaceted nature of political support across the nation's diverse populace. This paper seeks to dissect the nuanced ways in which factors such as race and income predict political support, thereby providing a more detailed understanding of American political behavior.

Our findings reveal significant correlations between sociodemographic characteristics and political support, highlighting the complexity of political engagement in the United States. These results not only highlight the importance of demographic factors in predicting political preferences but also illustrate the varied nature of political support across different segments of the American population. Additionally, the study identifies key issues, including economic policy and social justice, as crucial determinants of political support, offering new insights into the priorities and concerns that drive political behavior in the United States.

The paper begins with an overview of the CES 2020 dataset and the methods employed in our data analysis, followed by a presentation of results and key findings. We conclude with a discussion that contextualizes our results within the broader literature on political support in the United States. Our analysis was conducted using the statistical programming language R, utilizing packages such as tidyverse, dplyr, knitr, kableExtra, purrr, psych, car, MASS, and countrycode to facilitate data processing and visualization.

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

@@

install.packages("tidyverse")
install.packages("janitor")
install.packages("haven")
install.packages("dplyr")

library(tidyverse)
library(janitor) 
library(haven)
library(dplyr)

@@

# Breakdown of voter registration by race
registration_by_race <- cleaned_data %>% 
  group_by(race) %>% 
  summarise(
    Total = n(),
    Registered_Count = sum(votereg == "active", na.rm = TRUE),
    Registered_Percent = (sum(Registered_Count, na.rm = TRUE) / Total * 100
  )) %>% 
  arrange(desc(Registered_Percent))

print(registration_by_gender)

# Breakdown of voter registration by income
registration_by_income <- cleaned_data %>% 
  group_by(faminc_new) %>% 
  summarise(
    Total = n(),
    Registered_Count = sum(voter_status == "active", na.rm = TRUE),
    Registered_Percent = (sum(Registered_Count, na.rm = TRUE) / Total * 100
  )) %>% 
  arrange(desc(Registered_Percent))

print(registration_by_education)
```

@@

install.packages("tidyverse")
install.packages("janitor")
install.packages("haven")
install.packages("dplyr")
install.packages("ggplot2")

library(tidyverse)
library(janitor) 
library(haven)
library(dplyr)
library(ggplot2)

@@

install.packages("tidyverse")
install.packages("janitor")
install.packages("haven")
install.packages("dplyr")
install.packagess("ggplot2")

library(tidyverse)
library(janitor) 
library(haven)
library(dplyr)
library(ggplot2)

@@

ces2020 |>
  ggplot(data, aes(x = faminc_new, y = CC20_401, group = race, color = race)) +
  geom_jitter(width = 0.1, height = 0.1, alpha = 0.5) +
  stat_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  facet_wrap(~ race) +
  theme_minimal() +
  labs(x = "Income", y = "Biden", title = "Distribution of Voter Turnout by Race and Income") +
  scale_color_manual(values = c("Under $30000" = "blue", "$30000 to $49999" = "red", "$50000 to $99999" = "yellow", "$100000 to $149999" = "purple", "$200000 or more" = "green"))
```

@@

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

@@

# Model
To investigate the factors influencing voting behavior within the Cooperative Election Study (CES) 2020 dataset, we employed a logistic regression model, leveraging the `rstanarm` package of @rstanarm for its Bayesian inference capabilities. This choice allows for a more nuanced interpretation of results, incorporating prior beliefs and providing probabilistic estimates of the parameters. Secondly, this model aims to unpack the impact of demographic variables such as race and income on the probability of an individual casting a vote, framed within a binary outcome (voted vs. did not vote). The overarching goal of our modeling strategy is to distill the complex interplay of these demographic variables, helping our understanding of voting behaviors.

@@

### Model justification
For our model, we expect more minorities to vote for Biden based on his campaign. We also expect there to be a considerable effect of income. We intend to investigate if there is a significant relationship between these demographic factors and how they influence the probability of individuals voting for a particular candidate. To convey these relationships mathematically, we can represent them using LaTeX within dollar signs, such as $\beta_1$ for the effect of race and $\beta_2$ for the effect of income.

@@

library(rstanarm)

@@

political_preferences <-
  stan_glm(
    CC20_401 ~ race + faminc_new,
    data = ces2020,
    family = binomial(link = "logit"),
    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
    prior_intercept = 
      normal(location = 0, scale = 2.5, autoscale = TRUE)
  )

@@

modelsummary(
  list(
    "Support Biden" = political_preferences
  ),
  statistic = "mad"
  )
```

@@

income_levels <- data.frame(income = seq(min(ces2020$faminc_new, 
                                             na.rm = TRUE), 
                                        max(ces2020$faminc_new, 
                                            na.rm = TRUE), 
                                        length.out = 100))
# Predict probabilities
income_levels$prob_vote_biden <- predict(model, newdata = income_levels, 
                                         type = "response")

# Plot the probabilities
ggplot(income_levels, aes(x = faminc_new, y = prob_vote_biden)) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = 0, ymax = prob_vote_biden), fill = "blue", alpha = 0.2) +
  labs(title = "Predicted Probability of Voting for Biden by Income Level",
       x = "Income Level",
       y = "Probability of Voting for Biden") +
  theme_minimal()
```

@@

An important study in this context is Rhodes (2017), "Is America More Divided by Race or Class? Race, Income, and Attitudes among Whites, African Americans, and Latinos". Their findings suggest that race and class intersect in different ways for different groups in society. Increasing income erodes differences in attitudes between Latinos and whites, but has no effect on the large gap in attitudes between African Americans and whites. Interpolating these insights with the context of our model's outcomes, we gain a nuanced understanding of the interplay between culture, race, income, and voting behavior, emphasizing the critical role of cultural diversity in shaping political processes.

## Demographic Variable Effects {#sec-third-point}
We found that potentially due to varying policy priorities, historical voting trends, or the influence of race-specific issues,vVoter demographics can reveal how different racial groups may have distinct voting patterns. Analysis might show that income levels correlate with particular political preferences, where higher income brackets could favor policies or candidates that support their economic interests, while lower income brackets may support candidates advocating for social welfare programs.

@@

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

---
title: "Deciphering the Relationship Betweeen Race, Income, and Political Support in the 2020 United States Election"
author: 
  - Liban Timir
thanks: "Code and data are available at: https://github.com/libant/US-Political-Support-Analysis"
date: 03-12-2024
date-format: long
abstract: "In this paper, we delve into the intricate landscape of American politics using the comprehensive Cooperative Election Study (CES) of 2020, focusing on how race and income shape political support and preferences across the nation. Through a detailed analysis, I uncovered significant correlations between these socioeconomic factors and political allegiances, revealing the multifaceted nature of political engagement. Specifically, they both play crucial roles in influencing political behaviors and preferences, highlighting concerns that drive American political engagement. This research not only provides a deeper understanding of the dynamics at play in the current political climate but also emphasizes the importance of considering the varied perspectives and experiences within the American populace to grasp the full spectrum of political support in the United States."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In the contemporary political landscape of the United States, marked by increasing polarization, understanding the foundations of political support has never been more crucial. The Cooperative Election Study (CES) of 2020 [@ces2020] presents a dataset to explore this phenomenon. This dataset not only captures a wide array of sociopolitical attitudes and behaviors but also provides insights into the demographic and socioeconomic factors that influence political support. This offers a valuable lens through which to examine the complex dynamics of political engagement.

This study aims to fill a critical gap in the existing literature on political support in the United States by leveraging the CES 2020 data to examine the role of sociodemographic factors in shaping political preferences and behaviors. Previous research, while extensive, has often focused on more generalized measures of political engagement or on specific demographic groups, without fully addressing the multifaceted nature of political support across the nation's diverse populace. This paper seeks to dissect the nuanced ways in which factors such as race and income predict political support, thereby providing a more detailed understanding of American political behavior. The estimand is the correlations between socioeconomic factors such as income and political support.

The findings reveal significant correlations between these characteristics (income and race) and political support, highlighting the complexity of political engagement in the United States. These results not only highlight the importance of demographic factors in predicting political preferences but also illustrate the varied nature of political support across different segments of the American population, offering new insights into the priorities and concerns that drive political behavior in the United States.

The paper begins with an overview of the CES 2020 dataset and the methods employed in the data analysis, followed by a presentation of results and key findings. I conclude the paper with a discussion that contextualizes the results within the broader literature on political support in the United States. The analysis was conducted using the statistical programming language R [@citeR], utilizing packages such as tidyverse [@citetidyverse], dplyr [@citedplyr], ggplot2 [@citeggplot2], dataverse [@citedataverse], janitor [@citejanitor], rstanarm [@rstanarm], modelsummary [@modelsummary], broom.mixed [@broommixed] and additional papers such as Arceneaux's "Do Campaigns Help Voters Learn? A Cross-National Analysis" [@Arceneaux2005DoCH], Laurison's "Voting Intersections: Race, Class, and Participation in Presidential Elections in the United States 2008–2016" [@Laurison2021VotingIR], and Rhodes's "Is America More Divided by Race or Class? Race, Income, and Attitudes among Whites, African Americans, and Latinos"[@Rhodes2017IsAM].

@@

install.packages("tidyverse")
install.packages("janitor")
install.packages("dataverse")
install.packages("dplyr")

library(tidyverse)
library(janitor) 
library(dataverse)
library(dplyr)

@@

# Breakdown of voter registration by race
registration_by_race <- ces2020 %>% 
  group_by(race) %>% 
  summarise(
    Total = n(),
    Registered_Count = sum(votereg == "1", na.rm = TRUE),
    Registered_Percent = (sum(Registered_Count, na.rm = TRUE) / Total * 100
  )) %>% 
  arrange(desc(Registered_Percent))

print(registration_by_race)

# Breakdown of voter registration by income
registration_by_income <- ces2020 %>% 
  group_by(faminc_new) %>% 
  summarise(
    Total = n(),
    Registered_Count = sum(votereg == "1", na.rm = TRUE),
    Registered_Percent = (sum(Registered_Count, na.rm = TRUE) / Total * 100
  )) %>% 
  arrange(desc(Registered_Percent))

print(registration_by_income)
```

@@

install.packages("tidyverse")
install.packages("janitor")
install.packages("dataverse")
install.packages("dplyr")
install.packages("ggplot2")

library(tidyverse)
library(janitor) 
library(dataverse)
library(dplyr)
library(ggplot2)

@@

install.packages("tidyverse")
install.packages("janitor")
install.packages("dataverse")
install.packages("dplyr")
install.packages("ggplot2")

library(tidyverse)
library(janitor) 
library(dataverse)
library(dplyr)
library(ggplot2)

@@

ces2020 <- read_csv("ces2020.csv")
ces2020 <- ces2020 %>%
  mutate(race = as.factor(race))

ggplot(data = ces2020, aes(x = faminc_new, y = CC20_401, color = race)) +
  geom_jitter(width = 0.1, height = 0.1, alpha = 0.5) +
  stat_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, aes(group = race)) +
  facet_wrap(~ race) +
  theme_minimal() +
  labs(x = "Income", y = "Biden", title = "Distribution of Voter Turnout by Race and Income") +
  scale_color_manual(values = c("White" = "blue", "Black" = "red", "Hispanic" = "yellow", "Asian" = "purple", "Other" = "green"))
```

@@

@@

# Model
To investigate the factors influencing voting behavior within the Cooperative Election Study (CES) 2020 dataset, we employed a logistic regression model, leveraging the `rstanarm` package of @rstanarm for its Bayesian inference capabilities. This choice allows for a more nuanced interpretation of results, incorporating prior beliefs and providing probabilistic estimates of the parameters. Secondly, this model aims to unpack the impact of demographic variables such as race and income on the probability of an individual casting a vote, framed within a binary outcome (voted vs. did not vote). The overarching goal of the modeling strategy is to distill the complex interplay of these demographic variables, helping our understanding of voting behaviors.

@@

### Model justification
For the model, we expect more minorities to vote for Biden based on his campaign. We also expect there to be a considerable effect of income. We intend to investigate if there is a significant relationship between these demographic factors and how they influence the probability of individuals voting for a particular candidate. To convey these relationships mathematically, we can represent them using LaTeX within dollar signs, such as $\beta_1$ for the effect of race and $\beta_2$ for the effect of income.

@@

library(rstanarm)
library(dplyr)

@@

ces2020 <- read_csv("ces2020.csv")

ces2020 <- ces2020 %>%
  mutate(vote_recoded = case_when(
    CC20_410 == "Joe Biden (Democrat)" ~ 0,
    CC20_410 == "Donald J. Trump (Republican)" ~ 1,
    TRUE ~ NA_real_ 
  )) %>%
  na.omit() %>%
  select(vote_recoded, race, faminc_new)

political_preferences <-
  stan_glm(
    vote_recoded ~ race + faminc_new,
    data = ces2020,
    family = binomial(link = "logit"),
    prior = normal(0, 2.5),
    prior_intercept = normal(0, 2.5)
  )

@@

install.packages("rstanarm")
install.packages("modelsummary")
install.packages("broom.mixed")

library(rstanarm)
library(broom.mixed)
library(modelsummary)

if ("stan_glm" %in% class(political_preferences)) {
  tidy_summary <- tidy(political_preferences, parameters = TRUE)

  modelsummary(tidy_summary)
}
```

@@

biden_vote_proportion <- ces2020 %>%
  group_by(faminc_new) %>%
  summarise(Proportion_for_Biden = mean(CC20_401 == 1, na.rm = TRUE)) %>%
  ungroup()

ggplot(biden_vote_proportion, aes(x = faminc_new, y = Proportion_for_Biden)) +
  geom_col(fill = "blue") +  
  labs(title = "Proportion of Respondents Voting for Biden by Income",
       x = "Income",
       y = "Proportion Voting for Biden") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#income_levels <- data.frame(
#  income = seq(min(ces2020$faminc_new, na.rm = TRUE), 
#               max(ces2020$faminc_new, na.rm = TRUE), 
#               length.out = 100))

#if ("stan_glm" %in% class(political_preferences)) {
#  income_levels$prob_vote_biden <- predict(political_preferences, newdata = income_levels, type = "response")
#}

# Predict probabilities
#income_levels$prob_vote_biden <- predict(political_preferences, newdata = income_levels, type = "response")

# Plot the probabilities
#ggplot(income_levels, aes(x = income, y = prob_vote_biden)) +
#  geom_line(color = "blue") +
#  geom_ribbon(aes(ymin = 0, ymax = prob_vote_biden), fill = "blue", alpha = 0.2) +
#  labs(title = "Predicted Probability of Voting for Biden by Income Level",
#       x = "Income Level", 
#       y = "Probability of Voting for Biden") +
#  theme_minimal()
```

@@

An important study in this context is Rhodes (2017), "Is America More Divided by Race or Class? Race, Income, and Attitudes among Whites, African Americans, and Latinos". Their findings suggest that race and class intersect in different ways for different groups in society. Increasing income erodes differences in attitudes between Latinos and whites, but has no effect on the large gap in attitudes between African Americans and whites. Interpolating these insights with the context of the model's outcomes, we gain a nuanced understanding of the interplay between culture, race, income, and voting behavior, emphasizing the critical role of cultural diversity in shaping political processes.

## Demographic Variable Effects {#sec-third-point}
I found that potentially due to varying policy priorities, historical voting trends, or the influence of race-specific issues,vVoter demographics can reveal how different racial groups may have distinct voting patterns. Analysis might show that income levels correlate with particular political preferences, where higher income brackets could favor policies or candidates that support their economic interests, while lower income brackets may support candidates advocating for social welfare programs.

@@

#pp_check(first_model) +
#  theme_classic() +
#  theme(legend.position = "bottom")

#posterior_vs_prior(first_model) +
#  theme_minimal() +
#  scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
# coord_flip()
```

@@

#plot(first_model, "trace")

#plot(first_model, "rhat")
```
